A Note on Convergence Diagnostics for Multiple Imputation using Chained Equations (MICE)

### 2019-12-11 Simulation/Writeup/NoteConvergence.Rnw
Checked references

%% - \Abstract{} almost as usual
\Abstract{
The ubiquitous problem of missing information may be be circumvented by implementing the framework of multiple imputation (MI). With MI, missing data points are `filled in' (i.e. imputed) several times, and statistical inference is performed on each imputed data set. The validity of inferences may be threathened by non-convergence of this iterative algorithmic procedure, but diagnostics to evaluate algorithmic convergence have not been implemented in MI software. This note investigates the performance of two convergence diagnostics ($\widehat{R}$ \citep{gelm92}, auto-correlation XYZ) by means of model-based simulation. To investigate convergence of the algorithm, the number of iterations is varied between one and one-hundred. Each of these 100 simulation conditions is replicated 1000 times. The convergence diagnostics are evaluated against three recommended simulation diagnostics (bias, confidence interval width, and coverage rate \citep{buur18}), since no benchmark diagnostic is available. This simulation study shows that three iterations may be sufficient to obtain ... \textbf{This Research Report contains a simulation study that serves as the basis of the technical paper that will be submitted for publication in \emph{Journal of Statistical Software}. I aim to publish a pre-print of this research report on \href{https://arxiv.org/}{arXiv}. That way, I can refer to the simulation study described here, without `bulking up' the technical paper that I want to submit for publication in JSS. Another option would be to attach this note as online appendix to the technical paper on ShinyMICE.}  
% This Research Report contains a simulation study that serves as the basis of the technical paper that will be submitted for publication in \emph{Journal of Statistical Software}. I have chosen to use the first format from the Research Report guidelines: ``\emph{It is written as a (mini) thesis, with an introduction, methods section, some results (i.e., preliminary analyses, or pilot simulations), and a discussion of results. The length of the research report should be maximally 2500 words of text (without references list and or tables and figures). Please do not include appendices, and no more than 6 tables or figures. Table and Figure captions do not count towards the word limit. An abstract may be included, but is not necessary}''. I aim to publish a pre-print of this research report on \href{https://arxiv.org/}{ArXiv}. That way, I can refer to the simulation study described here, without 'bulking up' the technical paper that I want to submit for publication in JSS. Another option would be to attach this note as online appendix to the technical paper on ShinyMICE.
  }

@@

The goal of this research project is to develop novel methodology and guidelines for evaluating MI methods. These evaluation measures and guidelines will subsequently be implemented in an interactive evaluation framework for multiple imputation to aid applied researchers in drawing valid inference from incomplete datasets. This note provides the theoretical foundation for one of these evaluation measures that is vital in evaluting MI procedures: a diagnostic to assess convergence of the MI algorithm. More specifically, this note focusses on the MI algorithm that is implemented in the \proglang{R} package \pkg{mice}: `Multiple Imputation using Chained Equations' (MICE). The reseach question that will be adressed is: %. a measure that is vital in evaluating one vital measure for evaluating MI procedures--an algorithmic convergence diagnostic. This note specifically investigates one vital evaluation measure: how to diagnose convergence of a multiple imputation algorithm. 
%This note focusses on one measure that is vital for evaluating the validity of multiply imputed data: a convergence diagnostic. of the MI algorithm. 
%Therefore, the aim of this research project is to investigate which measures are vital for . how the convergence properties of the MI algorithm that is implemented in \pkg{mice}: `Multiple Imputation using Chained Equations' (MICE). This note addresses the question: 
\emph{`How to diagnose convergence of the multiple imputation algorithm MICE?'.} \textbf{REMOVE THE :'s HERE!}

The convergence properties of MICE are investigated by means of model-based simulation in \proglang{R} \citep{R}. The results of this simulation study are guidelines for assessing convergence of the MI algorithm. These guidelines will be implemented in an interactive evaluation tool for \pkg{mice}, `ShinyMICE', which is currently under development. All programming code used in this note is available from \href{https://github.com/gerkovink/shinyMice/simulation}{github.com/gerkovink/ShinyMICE/simulation}. 
%I will replicate Lacerda et al.'s simulation study on $\widehat{R}$ \citep{lace07}, and develop novel guidelines for assessing convergence. Ideally, I will integrate several diagnostics (e.g., $\widehat{R}$, \emph{auto-correlation}, and \emph{simulation error}) into a single summary indicator to flag non-convergence.

@@

The convergence guidelines introduced in this paper are developed to be integrated into the \pkg{mice} environment \citep{mice} in \proglang{R} \citep{R}. This note therefore follows notation and conventions of \cite{mice}. Deviations from the `original' notation by \cite{rubin87} are described in \citep[\S~2.2.3]{buur18}.

@@

This note focusses on the algorithmic properties of the imputation step. The algorithm employed within this step has an iterative nature. That is, before drawing $m$ imputed values for each missing data point in $Y_{mis}$, a `chain' of potential values is sampled. Each of the $m$ chains starts with an initial value, drawn randomly from $Y_{obs}$. The chains are terminated after a predefined number of iterations. Only the ultimate sample that a chain lands on is imputed, and subsequently used in the next steps (analysis and pooling). The collection of samples across iterations (between the initial value and the imputed value) for each of the $m$ imputations will be referred to as an `imputation chain'. The number of iterations will be denoted with $t$.

@@

The multiple imputation algorithm `MICE' is a special case of Markov chain Monte Carlo (MCMC) methods--the general framework of iterative algorithmic procedures. While methodology for diagnosing convergence of multiple imputation algorithms has not been systematically studied \citep{buur18}, several convergence diagnostics have been proposed for other types of MCMC methods (e.g., Gibbs samplers), or the framework in general.
% For the framework of itrIt is not known whether convergence diagnostics that are available for the general framework that MICE is a special case of are appropriate for the evaluation of MI data. This is the framework of Markoc chain Monte Carlo %Yet, reference manuals of promiment MI software do mention the importance of evaluating algorithmic convergence (e.g., \cite[p.XYZ]{mice}, \textbf{XYZ, XYZ}). 
This section provides a brief review of: 1) how convergence is defined in the general context of iterative algorithmic procedures; 2) what diagnostic tools are available; and 3) which of these may be applicable to MI algorithms.

Losely speaking, the aim of MCMC algorithms (e.g., Gibbs samplers, or MICE) is to sample values from an unknown target distribution (e.g., the posterior distribution of a parameter in Bayesian inference, or a distribution of missing values, respectively). Convergence is reached once the algorithm samples exclusively from the target distribution \citep{cowl96}. Since this target distribution is unknown by definition, convergence can only be monitored by evaluating signs of non-convergence \citep{hoff09}. Therefore, in practice, the definition of convergence consists of two components: `mixing' of chains together, and `stationarity' of individual chains \citep[p.~284]{gelm13}. The mixing component is satisfied when chains intermingle such that the only difference between the chains is caused by the randomness induced by the algorithm. Stationarity refers to the absence of trending within chains. \textbf{Figure XYZ shows a pathological case of non-convergence with respect to both components.}  
%The aim of MCMC procedures is to sample from an unknown (theoretical) distribution. The stability of the solution denpends on two components of convergence: should have converged, two components should be satisfied.

@@

The two components of convergence may be evaluated with convergence diagnostics. Diagnostics that target the stationarity component are auto-correlation \citep{gelm13}, numeric standard error \citep[which is commonly called `MC error']{gewe92}, and Raftery and Lewis's \citeyear{raft92} procedure to determine the increase in the necessary number of iterations due to dependence within chains. The mixing component of convergence, on the other hand, is primarily assessed with the `Gelman-Rubin statistic'--potential scale reduction factor $\widehat{R}$ \citep{gelm92}. An adapted version of $\widehat{R}$ has been proposed by \cite{veht19}, which may also assess the stationarity component of convergence. Therefore, the focus of this study is mainly on $\widehat{R}$. Stationarity will also be evaluated with auto-correlation as diagnostic, as recommended by \cite[p.~898]{cowl96}. Other convergence diagnostics are outside the scope of this study. \footnotemark 
\footnotetext{All of these methods evaluate the convergence of univariate scalar summaries (e.g., chain means or variances). These convergence diagnostics cannot diagnose convergence of multivariable statistics (i.e., relations between scalar summaries). \cite{buur18} proposed to implement multivariable evaluation by means of eigenvalue decomposition \cite{mack03}. This method is outside of the scope of the current study.}%  he eigenvector decomposition method proposed by McKay \citep{XYZ}.} These four methods, however, are outside of the scope of this study. Here, the focus is mainly on $\widehat{R}$ and auto-correlation.

@@

It is not known whether any convergence diagnostic is appropriate to diagnose convergence of the MICE algorithm. As recommended by \cite[\S~6.5.2]{buur18}, the current practice is to visually inspect imputation chains for signs of non-mixing and non-stationarity. Not only is visually assessing convergence challenging to the untrained eye, it might also be futile. That is, there is no scientific consensus on the convergence properties of multiple imputation algorithms \citep{taka17}, and in some situations the default techniques in \pkg{mice} might not converge to stable target distributions at all \citep{murr18}. %Convergence diagnostics for MI methods have not been systematically studied \citep{buur18}.  Moreover, 
%A fundamental assumption of MICE is convergence of the algorithm. MICE is a type of Markov chain Monte Carlo (MCMC) algorithm. In general, the validity of inference resulting from MCMC algorithms is threatened by non-convergence. Hence we use convergence diagnostics to flag non-convergence. But it is not known whether conventional convergence diagnostics for MCMC methods work on MI data.
Therefore, the need for convergence diagnostics is extra pressing in the context of the MICE algorithm.

@@

\textbf{Explain the measures that I'm investigating here. R hat is ... Auto-correlation is... How are they computed? What do they say? How are they computed on MI data? Looking at imputation chains. 
Explain that there is no baseline to compare the measures with. Only current convergence check is to visually inspect trace plots. (But in the trace-plots convergence seems immediate).}

The potential scale reduction factor $\widehat{R}$ tells us by how much the variance of an estimate could be shrunken down, if the chains were to be infinitely long \cite{gelm92}. That then informs us about how dependent the chains are on the starting values. If there is no dependence on the initial values anymore, the chains have converged (in the mixing sense of the word). $\widehat{R}$ is then equal to one. The conventional acceptable threshold was $\widehat{R} < 1.2$. More recently, \cite{veht19} proposed a more stringent threshold: $\widehat{R} < 1.01$.

@@

The second component of convergence stationarity requires chains to be stable across iterations, or non-recurring. Non-recurrence can be evaluated with auto-correlation. Auto-correlation shows how dependent subsequent draws of an imputation chain are on the previous value. If there is a lot of dependence, draws at e.g. iteration five are significantly correlated with the value of the first draw. A high auto-correlation indicates dependence within chains. The magnitude of the AC can be interpreted qualitatively, or compared to the critical value of a \textbf{XYZ} distribution with \textbf{XYZ} degrees of freedom.

@@

The hypothesis of this simulation study is that $\widehat{R}$ will over-estimate non-convergence. \textbf{[Or use:] Hypothesis based on \cite{lace07} is that the conventional acceptable level of $\widehat{R}$ is too strict for MI data. We expect that the simulation diagnostics will indicate valid inference before $\widehat{R}$ will.}

@@

To investigate convergence properties of the \pkg{mice} algorithm, different lenghts of imputation chains are considered. This simulation study considers 100 simulation conditions. Each condition has a different maximum number of iterations, varying between one and one-hundred iterations. In this study, 1000 simulation runs are performed for each of the 100 simulation conditions. \textbf{where for one through one hundred the Rhat and ... are calcuated.} The simulation set-up consists of several steps, summarized in the pseudo-code below. The complete \proglang{R} script of the simulation study is available on Github.

@@

\subsection{Imputation}
Missing data points are imputed with \pkg{mice} in \proglang{R}. All simulations are performed with imputation method 'norm' (Bayesian linear regression imputation), and five imputation chains ($m = 5$). The number of iterations is varied over simulations ('maxit' argument between 1 and 100). Each simulation condition (i.e., each number of iterations) is simulated 1000 times. Simulation and convergence diagnostics are aggregated over the 1000 MCMC simulations. 
% Missingness mechanism: "With MCAR missingness mechanisms, the probability to be missing is the same for all cases. This is a necessary simulation condition for evaluating the performance of imputation procedures. If an imputation method is not able to solve the problem (i.e. yield valid inference) under MCAR, the statistical properties of the procedure are not sound." (Vink, n.d., p. 4)
% "combine the m completed data inferences into a single inference following Rubin's (1987) rules (pp. 76-77)" \cite{schou18}.

@@

Auto-correlation (AC) is computed within imputation chains, as the correlation between $\theta$ at the $i^{th}$ iteration and $\theta$ at the $i+1^{th}$ iteration, where $\theta$ is a scalar summary of interest (i.e., chain means and chain variances). The AC of the variable with the highest absolute AC value is reported per simulation. These values are then averaged per simulation condition.

@@

\textbf{$\widehat{R}$}. The $\widehat{R}$ values for chain means and chain variances show equivalent trends, neither is there a clear difference between the four variables. Roughly speaking, we see that $\widehat{R}$ is higher for simulation conditions with a lower number of iterations, and approaches one for conditions with more iterations. Figure \ref{fig:conv} shows an initial dip, a small bump upwards, and then a steep decline. This decline \textbf{houdt aan} up-to iteration twenty, after which a more gradual decrease can be observed up-to iteration forty. The $\widehat{R}$ values are more or less stable after that. The conventional acceptable threshold for convergence, maximum $\widehat{R} < 1.2$, is reached in the simulation condition with three iterations, and conditions with six iterations or more. The widely used threshold $\widehat{R} < 1.1$ is met in conditions with ten or more iterations. The most recent recommended threshold is not reached within the 100 iterations considered in this simulation study. $\widehat{R} > 1.01$ for all simulation conditions (see online appendix for the full table of results). \textbf{Check is computing r ht is still correct!}

\textbf{Auto-correlation.} All observed auto-correlations (aggregated over the 1000 repetitions) are negative or close to zero. There are some differences between variables, and the ACs of chain variances seem lower than the ACs of chain means. The general trend in figure \ref{fig:conv} is that ACs decrease steeply as the number of iterations increase.  \textbf{The lowest AC observed is %\Sexpr{round(min(dat[-1, c("AC chain mean", "AC chain var")]), 3)} 
at four iterations. Simulation conditions with a higher number of iterations have show a slow increase towards zero. Auto-correlations do not plateau completely within the range of simulation conditions considered in this study. It seems that ACs might become positive if more iterations would be allowed. The average absolute maximum ACs are negative up to 77 iterations for chain means, and for chain variances all iterations considered here have negative ACs.}

@@

%% --------------------------------
\subsection{Simulation diagnostics}
\textbf{Bias.} Figure \ref{fig:sim} shows the average bias between the regression coefficient of $X$ on $Y$ in the complete dataset, and $\bar{Q}$, \textbf{the pooled estimated regression coefficient and the  is close to zero for simulation conditions with two iterations or more. The simulation condition with one iteration has a negative bias. From two iterations upwards, the average bias across repetitions is stable. Across iterations two to one hundred, bias fluctuates within a narrow range around zero. Loess line is flat around it = 20.} 
% Average bias is not zero because there is only one incomplete dataset to be imputed. Bias is a sample effect.

@@

\textbf{Summary.}
This note illustrates that conventional convergence diagnostics behave differently on MI data than other MCMC methods like Gibbs samplers in Bayesian analyses. The most recent recommended threshold for $\widehat{R}$ ($\widehat{R} < 1.01$) may be too stringent for MI data.

From the simulation diagnostics, it appears that as little as two iterations could be sufficient to draw \textbf{[valid inferences?]}. Convergence diagnostics $\widehat{R}$ and auto-correlation, however, indicate that convergence may only be reached after twenty or even forty iterations.

@@

\textbf{Future research.} %We would like to have convergence measures for multivariable statistics (scalars?) of interest. This is, however, dependent on the complete data model. The eigenvector decomposition method proposed by McKay (?) should be implemented. I could not find any resources to apply this method and it is outside the scope of this thesis to investigate how this approach could be implemented. \textbf{[this paragraph is too negative, explain scope in intro instead.]}

>>>>

%% - \Abstract{} almost as usual
\Abstract{
The ubiquitous problem of missing information may be circumvented by implementing the framework of multiple imputation (MI). With MI, missing data points are `filled in' (i.e. imputed) several times, and statistical inference is performed on each imputed data set. The validity of inferences may be threatened by non-convergence of this iterative algorithmic procedure, but diagnostics to evaluate algorithmic convergence have not been implemented in MI software. This note investigates the performance of two convergence diagnostics ($\widehat{R}$ \citep{gelm92}, auto-correlation \citep{scha97}) through model-based simulation. To investigate convergence of the algorithm, the number of iterations is varied between one and one-hundred. Each of these 100 simulation conditions is replicated 1000 times. The convergence diagnostics are evaluated against three recommended simulation diagnostics (bias, confidence interval width, and coverage rate \citep{buur18}, since no benchmark diagnostic is available. This simulation study shows that three iterations may be sufficient to obtain ... \textbf{This Research Report contains a simulation study that serves as the basis of the technical paper that will be submitted for publication in \emph{Journal of Statistical Software}. I aim to publish a pre-print of this research report on \href{https://arxiv.org/}{arXiv}. That way, I can refer to the simulation study described here, without `bulking up' the technical paper that I want to submit for publication in JSS. Another option would be to attach this note as an online appendix to the technical paper on ShinyMICE.}  
% This Research Report contains a simulation study that serves as the basis of the technical paper that will be submitted for publication in \emph{Journal of Statistical Software}. I have chosen to use the first format from the Research Report guidelines: ``\emph{It is written as a (mini) thesis, with an introduction, methods section, some results (i.e., preliminary analyses, or pilot simulations), and a discussion of results. The length of the research report should be maximally 2500 words of text (without references list and or tables and figures). Please do not include appendices, and no more than 6 tables or figures. Table and Figure captions do not count towards the word limit. An abstract may be included, but is not necessary}''. I aim to publish a pre-print of this research report on \href{https://arxiv.org/}{ArXiv}. That way, I can refer to the simulation study described here, without 'bulking up' the technical paper that I want to submit for publication in JSS. Another option would be to attach this note as online appendix to the technical paper on ShinyMICE.
  }

@@

The goal of this research project is to develop novel methodology and guidelines for evaluating MI methods. These evaluation measures and guidelines will subsequently be implemented in an interactive evaluation framework for multiple imputation to aid applied researchers in drawing valid inference from incomplete datasets. This note provides the theoretical foundation for one of these evaluation measures that is vital in evaluating MI procedures: a diagnostic to assess convergence of the MI algorithm. More specifically, this note focuses on the MI algorithm that is implemented in the \proglang{R} package \pkg{mice}: `Multiple Imputation using Chained Equations' (MICE). The research question that will be addressed is: %. a measure that is vital in evaluating one vital measure for evaluating MI procedures--an algorithmic convergence diagnostic. This note specifically investigates one vital evaluation measure: how to diagnose convergence of a multiple imputation algorithm. 
%This note focusses on one measure that is vital for evaluating the validity of multiply imputed data: a convergence diagnostic. of the MI algorithm. 
%Therefore, the aim of this research project is to investigate which measures are vital for . how the convergence properties of the MI algorithm that is implemented in \pkg{mice}: `Multiple Imputation using Chained Equations' (MICE). This note addresses the question: 
\emph{`How to diagnose convergence of the multiple imputation algorithm MICE?'.} \textbf{REMOVE THE :'s HERE!}

The convergence properties of MICE are investigated through model-based simulation in \proglang{R} \citep{R}. The results of this simulation study are guidelines for assessing convergence of the MI algorithm. These guidelines will be implemented in an interactive evaluation tool for \pkg{mice}, `ShinyMICE', which is currently under development. All programming code used in this note is available from \href{https://github.com/gerkovink/shinyMice/simulation}{github.com/gerkovink/ShinyMICE/simulation}. 
%I will replicate Lacerda et al.'s simulation study on $\widehat{R}$ \citep{lace07}, and develop novel guidelines for assessing convergence. Ideally, I will integrate several diagnostics (e.g., $\widehat{R}$, \emph{auto-correlation}, and \emph{simulation error}) into a single summary indicator to flag non-convergence.

@@

The convergence guidelines introduced in this paper are developed to be integrated into the \pkg{mice} environment \citep{mice} in \proglang{R} \citep{R}. This note, therefore, follows notation and conventions of \cite{mice}. Deviations from the `original' notation by \cite{rubin87} are described in \citep[\S~2.2.3]{buur18}.

@@

This note focuses on the algorithmic properties of the imputation step. The algorithm employed within this step has an iterative nature. That is, before drawing $m$ imputed values for each missing data point in $Y_{mis}$, a `chain' of potential values is sampled. Each of the $m$ chains starts with an initial value, drawn randomly from $Y_{obs}$. The chains are terminated after a predefined number of iterations. Only the ultimate sample that a chain lands on is imputed and subsequently used in the analysis and pooling steps. The collection of samples across iterations (between the initial value and the imputed value) for each of the $m$ imputations will be referred to as an `imputation chain'. The number of iterations will be denoted with $t$ \textbf{[NOG TOEPASSEN]}.

@@

The multiple imputation algorithm `MICE' is a special case of Markov chain Monte Carlo (MCMC) methods--the general framework of iterative algorithmic procedures. While methodology for diagnosing convergence of multiple imputation algorithms has not been systematically studied \citep{buur18}, several convergence diagnostics have been proposed for other types of MCMC methods (e.g., Gibbs samplers), or the framework in general.
% For the framework of itrIt is not known whether convergence diagnostics that are available for the general framework that MICE is a special case of are appropriate for the evaluation of MI data. This is the framework of Markoc chain Monte Carlo %Yet, reference manuals of promiment MI software do mention the importance of evaluating algorithmic convergence (e.g., \cite[p.XYZ]{mice}, \textbf{XYZ, XYZ}). 
This section provides a brief review of: 1) how convergence is defined in the general context of iterative algorithmic procedures; 2) what diagnostic tools are available; and 3) which of these may apply to MI algorithms.

Loosely speaking, the aim of MCMC algorithms (e.g., Gibbs samplers, or MICE) is to sample values from an unknown target distribution (e.g., the posterior distribution of a parameter in Bayesian inference, or a distribution of missing values, respectively). Convergence is reached once the algorithm samples exclusively from the target distribution \citep{cowl96}. Since this target distribution is unknown by definition, convergence can only be monitored by evaluating signs of non-convergence \citep{hoff09}. Therefore, in practice, the definition of convergence consists of two components: `mixing' of chains together, and `stationarity' of individual chains \citep[p.~284]{gelm13}. The mixing component is satisfied when chains intermingle such that the only difference between the chains is caused by the randomness induced by the algorithm. Stationarity refers to the absence of trending within chains. \textbf{Figure XYZ shows a pathological case of non-convergence concerning both components.}  
%The aim of MCMC procedures is to sample from an unknown (theoretical) distribution. The stability of the solution denpends on two components of convergence: should have converged, two components should be satisfied.

@@

The two components of convergence may be evaluated with convergence diagnostics. Diagnostics that target the stationarity component are auto-correlation \citep{scha97, gelm13}, numeric standard error \citep[which is commonly called `MC error']{gewe92}, and Raftery and Lewis's \citeyear{raft91} procedure to determine the increase in the necessary number of iterations due to dependence within chains. The mixing component of convergence, on the other hand, is primarily assessed with the `Gelman-Rubin statistic'--potential scale reduction factor $\widehat{R}$ \citep{gelm92}. An adapted version of $\widehat{R}$ has been proposed by \cite{veht19}, which may also assess the stationarity component of convergence. Therefore, the focus of this study is mainly on $\widehat{R}$. Stationarity will also be evaluated with auto-correlation as diagnostic, as recommended by \cite[p.~898]{cowl96}. Other convergence diagnostics are outside the scope of this study. \footnotemark 
\footnotetext{All of these methods evaluate the convergence of univariate scalar summaries (e.g., chain means or variances). These convergence diagnostics cannot diagnose convergence of multivariable statistics (i.e., relations between scalar summaries). \cite{buur18} proposed to implement multivariable evaluation by means of eigenvalue decomposition \cite{mack03}. This method is outside of the scope of the current study.}%  he eigenvector decomposition method proposed by McKay \citep{XYZ}.} These four methods, however, are outside of the scope of this study. Here, the focus is mainly on $\widehat{R}$ and auto-correlation.

@@

It is not known whether any convergence diagnostic is appropriate to diagnose convergence of the MICE algorithm. As recommended by \cite[\S~6.5.2]{buur18}, the current practice is to visually inspect imputation chains for signs of non-mixing and non-stationarity. Not only is visually assessing convergence challenging to the untrained eye--it might also be futile. That is, there is no scientific consensus on the convergence properties of multiple imputation algorithms \citep{taka17}, and in some situations, the default techniques in \pkg{mice} might not converge to stable target distributions at all \citep{murr18}. %Convergence diagnostics for MI methods have not been systematically studied \citep{buur18}.  Moreover, 
%A fundamental assumption of MICE is convergence of the algorithm. MICE is a type of Markov chain Monte Carlo (MCMC) algorithm. In general, the validity of inference resulting from MCMC algorithms is threatened by non-convergence. Hence we use convergence diagnostics to flag non-convergence. But it is not known whether conventional convergence diagnostics for MCMC methods work on MI data.
Therefore, the need for convergence diagnostics is extra pressing in the context of the MICE algorithm.

@@

\textbf{Explain the measures that I'm investigating here. R hat is ... Auto-correlation is... How are they computed? What do they say? How are they computed on MI data? Looking at imputation chains. 
Explain that there is no baseline to compare the measures with. The only current convergence check is to visually inspect trace plots. (But in the trace-plots convergence seems immediate).}

The potential scale reduction factor $\widehat{R}$ tells us by how much the variance of an estimate could be shrunken down if the chains were to be infinitely long \cite{gelm92}. That then informs us about how dependent the chains are on the starting values. If there is no dependence on the initial values anymore, the chains have converged (in the mixing sense of the word). $\widehat{R}$ is then equal to one. The conventionally acceptable threshold was $\widehat{R} < 1.2$. More recently, \cite{veht19} proposed a more stringent threshold: $\widehat{R} < 1.01$.

@@

The second component of convergence stationarity requires chains to be stable across iterations, or non-recurring. Non-recurrence can be evaluated with auto-correlation. Auto-correlation shows how dependent subsequent draws of an imputation chain are on the previous value. If there is a lot of dependence, draws at e.g. iteration five are significantly correlated with the value of the first draw. A high auto-correlation indicates dependence within chains. The magnitude of the AC can be interpreted qualitatively or quantitatively. Quantitative evaluation of the AC entails comparing the observed ACs to the critical values of a two-tailed 95\% confidence interval, divided by the square root of the number of iterations. %. critical value corresponding to the $z-$~value standard normal distribution with \textbf{XYZ} degrees of freedom.
%  AC_lim <- qnorm(.975) / sqrt(maxit)

@@

This simulation study hypothesizes that $\widehat{R}$ will over-estimate non-convergence. \textbf{[Or use:] Hypothesis based on \cite{lace07} is that the conventional acceptable level of $\widehat{R}$ is too strict for MI data. We expect that the simulation diagnostics will indicate valid inference before $\widehat{R}$ will.}

@@

To investigate the convergence properties of the \pkg{mice} algorithm, different lengths of imputation chains are considered. This simulation study considers 100 simulation conditions. Each condition has a different maximum number of iterations, varying between one and one-hundred iterations. In this study, 1000 simulation runs are performed for each of the 100 simulation conditions. \textbf{Estimands are convergence diagnostics. Performance measures are simulation diagnostics.  where for one through one hundred the Rhat and ... are calculated.} The simulation set-up consists of several steps, summarized in the pseudo-code below. The complete \proglang{R} script of the simulation study is available on Github.

@@

\subsection{Imputation}
Missing data points are imputed with \pkg{mice} in \proglang{R}. All simulations are performed with imputation method 'norm' (Bayesian linear regression imputation), and five imputation chains ($m = 5$). The number of iterations is varied over simulations (\texttt{maxit} argument between 1 and 100). Each simulation condition (i.e., each number of iterations) is simulated 1000 times. Simulation and convergence diagnostics are aggregated over the 1000 MCMC simulations. 
% Missingness mechanism: "With MCAR missingness mechanisms, the probability to be missing is the same for all cases. This is a necessary simulation condition for evaluating the performance of imputation procedures. If an imputation method is not able to solve the problem (i.e. yield valid inference) under MCAR, the statistical properties of the procedure are not sound." (Vink, n.d., p. 4)
% "combine the m completed data inferences into a single inference following Rubin's (1987) rules (pp. 76-77)" \cite{schou18}.

@@

Auto-correlation (AC) is computed within imputation chains, as the correlation between $\theta$ at \textbf{the $i^{th}$ iteration and $\theta$ at the $i+1^{th}$ iteration [CHECK VARIABLE IN NOTATION]}, where $\theta$ is a scalar summary of interest (i.e., chain means and chain variances). The AC of the variable with the highest absolute AC value is reported per simulation. These values are then averaged per simulation condition.

@@

\textbf{$\widehat{R}$}. The $\widehat{R}$ values for chain means and chain variances show equivalent trends, neither is there a clear difference between the four variables. Roughly speaking, we see that $\widehat{R}$ is higher for simulation conditions with a lower number of iterations, and approaches one for conditions with more iterations. Figure \ref{fig:conv} shows an initial dip, a small bump upwards, and then a steep decline. This decline \textbf{houdt aan} up-to iteration twenty, after which a more gradual decrease can be observed up-to iteration forty. The $\widehat{R}$ values are more or less stable after that. The \textbf{conventionally acceptable threshold for convergence, maximum $\widehat{R} < 1.2$,[CHECK IN INTRO]} is reached in the simulation condition with three iterations, and conditions with six iterations or more. The widely used threshold $\widehat{R} < 1.1$ is met in conditions with ten or more iterations. The most recent recommended threshold is not reached within the 100 iterations considered in this simulation study. $\widehat{R} > 1.01$ for all simulation conditions (see online appendix for the full table of results). \textbf{Check is computing r ht is still correct! Also say something about Rh below 1.}

\textbf{Auto-correlation.} All observed auto-correlations (aggregated over the 1000 repetitions) are negative or close to zero. There are some differences between variables, and the ACs of chain variances seem lower than the ACs of chain means. The general trend in figure \ref{fig:conv} is that ACs decrease steeply as the number of iterations increases.  \textbf{The lowest AC observed is %\Sexpr{round(min(dat[-1, c("AC chain mean", "AC chain var")]), 3)} 
at four iterations. Simulation conditions with a higher number of iterations show a slow increase towards zero. Auto-correlations do not plateau completely within the range of simulation conditions considered in this study. It seems that ACs might become positive if more iterations would be allowed. The average absolute maximum ACs are negative up to 77 iterations for chain means, and for chain variances all iterations considered here have negative ACs.}

@@

%% --------------------------------
\subsection{Simulation diagnostics}
\textbf{Bias.} Figure \ref{fig:sim} shows the average bias between the regression coefficient of $X$ on $Y$ in the complete dataset, and $\bar{Q}$, \textbf{the pooled estimated regression coefficient and the ... Average bias is close to zero for simulation conditions with two iterations or more. The simulation condition with one iteration has a negative bias. From two iterations upwards, the average bias across repetitions is stable. Across iterations two to one hundred, bias fluctuates within a narrow range around zero. Loess line is flat around it = 20.} 
% Average bias is not zero because there is only one incomplete dataset to be imputed. Bias is a sample effect.

@@

\textbf{Summary.}
This note illustrates that conventional convergence diagnostics behave differently on MI data than other MCMC methods like Gibbs samplers in Bayesian analyses. The most recent recommended threshold for $\widehat{R}$ ($\widehat{R} < 1.01$) may be too stringent for MI data. \textbf{However, using the 1.2 threshold may be to lenient because this was observed at three iterations and then again after six. If we would use 1.2, we might miss this increase after three iterations. The 1.1 threshold seems ok. Also, ten iterations are computationally not terrible. With increased performance and storage capacities, doubling the default is fine I guess.}

From the simulation diagnostics, it appears that as little as three iterations might be sufficient to obtain unbiased, confidence valid estimates. Convergence diagnostics $\widehat{R}$ and auto-correlation, however, indicate that convergence may only be reached after twenty or even forty iterations.

@@

\textbf{Future research.} This study consdered only a MCAR missingness mechanism. Necessary but not sufficient to demonstrate the performance of convergence diagnostics. This is just a proof of concept. Further reseach is needed to investigate the performance under violation of convergence, e.g. dependency between predictors (very high correlations). %We would like to have convergence measures for multivariable statistics (scalars?) of interest. This is, however, dependent on the complete data model. The eigenvector decomposition method proposed by McKay (?) should be implemented. I could not find any resources to apply this method and it is outside the scope of this thesis to investigate how this approach could be implemented. \textbf{[this paragraph is too negative, explain scope in intro instead.]}
